---
title: "Using skrmdb"
author: "Thomas Kent"
date: "`r Sys.Date()`"
output:  html_vignette
vignette: >
  %\VignetteIndexEntry{Using skrmdb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo=FALSE}
library(skrmdb)
library(ggplot2)
```

# Notation
The methods of Spearman-Karber, Reed-Muench, and Dragstedt-Behrens are all
commonly used to estimate ED50. In what follows, we use the following variable
notation:

-   `x` is a vector corresponding to the log dilution or dose for each group.
-   `n` is an integer vector corresponding to the group size at each log
dilution or dose.
-   `y` is an integer vector corresponding to the number responding at each log
dilution or dose.

All examples are with `SpearKarb()`, however, the usage for `DragBehr()` and
`ReedMuench()` is identical.

# Usage

Each of the main functions in `skrmdb` can be called in three different ways.

To illustrate this, we start with a simple data set where the number of dead
increases with the dosage.

```{r}
dead <- c(0, 3, 5, 8, 10, 10)
total <- rep(10, 6)
dil <- 1:6
```

First we use the historical, deprecated, function call.

```{r}
SpearKarb(y = dead, n = total, x = dil)
```

The preferred method is to use formulas in the function call.

We can use the formula on the individual vectors.

```{r}
SpearKarb(dead + total ~ dil)
```

We can also use the formula on columns in a data.frame.

```{r}
data <- data.frame(y = dead, n = total, x = dil)
SpearKarb(y + n ~ x, data)
```

# Method Assumptions

Each of these methods was designed to work under the assumption that x is either 
increasing or decreasing and y / n is increasing with the index.

It was decided that the functions `DragBehr()`, `ReedMuench()`, and
`SpearKarb()` will sort the data for you according to this assumption.

To illustrate this, we create some data that is descending according to dilution.

```{r}
dead <- c(10, 10, 8, 5, 3, 0)
total <- rep(10, 6)
dil <- 1:6
```

And we see that no mater how we enter the data, the same ED50 is reported.

```{r}
SpearKarb(dead + total ~ dil)
SpearKarb(rev(dead) + rev(total) ~ rev(dil))
```

However, if we can decided to turn off the autosort, which will give us a
different, but incorrect, estimate of ED50.

```{r}
SpearKarb(dead + total ~ dil, autosort = FALSE)
```

# Conditional ED50

The function `skrmdb.all` returns the ED50 for the three methods, along with
additional data about the data sets that may be of interest.

For this section, we use the example data set `titration`. This `data.frame`
contains the results of a hypothetical experiment where the ED50 of three vials
(numbered 1, 2, 3) were each tested by three anonymous operators (TK, NU, CT).

```{r}
head(titration)
```

First we need to compute the log dilution.

```{r}
titration$log_dil <- -log10(titration$dil)
```

Plotting the data shows us that the tests visually seem to give roughly the 
same ED50, but unfortunately, there are missing data points.

```{r, fig.alt="Nine plots showing titration data."}
ggplot(titration, aes(x = log_dil, y = positive)) +
  geom_point() +
  facet_grid(Vial ~ Operator) +
  geom_line()
```

We could find ED50 by aggregation the data from all 9 tests for each of the
three methods.

```{r}
skrmdb.all(positive + total ~ log_dil, titration)
```

We can also see if there is a difference between Operators;

```{r}
skrmdb.all(positive + total ~ log_dil | Operator, titration)
```

Or by Vial;

```{r}
skrmdb.all(positive + total ~ log_dil | Vial, titration)
```

Or by Operator and Vial.

```{r}
skrmdb.all(positive + total ~ log_dil | Vial + Operator, titration)
```

# Additional Methods

Finally, there are three accessor functions to help retrieve information from
the results of `DragBehr()`, `ReedMuench()`, and `SpearKarb()`.

Using the titration data set again.

```{r}
res <- SpearKarb(positive + total ~ log_dil, titration)
```

We can get the ED50:  (**Note:** `getED50()` only works on `skrmdb` objects, 
which are returned by `DragBehr()`, `ReedMuench()`, and `SpearKarb()`,
not on `skrmdb.all` objects.)

```{r}
getED50(res)
```

The variance:

```{r}
getvar(res)
```

And the data which was used.

```{r}
getdata(res)
```
