---
title: "Using skrmdb"
author: "Center for Veterinary Biologics - Statistics Section"
date: "`r Sys.Date()`"
output:  html_vignette
vignette: >
  %\VignetteIndexEntry{Using skrmdb}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---


```{r}
#| echo: false
library(skrmdb)
library(ggplot2)
```

# Notation

The methods of Spearman-Kärber, Reed-Muench, and Dragstedt-Behrens are all commonly used to estimate ED50. In what follows, we use the following variable notation:

-   `x` is a vector corresponding to the log dilution or dose for each group.
-   `n` is an integer vector corresponding to the group size at each log dilution or dose.
-   `y` is an integer vector corresponding to the number responding at each log dilution or dose.

All examples use `SpearKarb()`. However, the usage for `DragBehr()`, `ReedMuench()`, and `skrmdb.all()` is identical.

# Usage

Each of the main functions in `{skrmdb}` can be called in three different ways.

To illustrate this, we start with a simple data set where the number dead increases with the dosage.

```{r}
dead <- c(0, 3, 5, 8, 10, 10)
total <- rep(10, 6)
dil <- 1:6
```

## Example 1

First we use the deprecated historical function call.  We will also suppress extra messages that help interpret the data.

```{r}
print(
  SpearKarb(y = dead, n = total, x = dil),
  verbose = FALSE
)

```

## Example 2

The preferred method is to use formulas in the function call.

We can use a formula with individual vectors.

```{r}
#| message: false
print(
  SpearKarb(dead + total ~ dil),
  verbose = FALSE
)

```

## Example 3

We can also use the same formula on columns in a `data.frame`.

```{r}
#| message: false
data <- data.frame(y = dead, n = total, x = dil)
print(
  SpearKarb(data, y + n ~ x),
  verbose = FALSE
)

```

# Conditional ED50

ED50 can also be estimated conditional on grouping variables.  To illustrate this, we use the example data set `titration`. This `data.frame` contains the results of a hypothetical experiment where the ED50 in which three vials (numbered 1, 2, 3) were each tested by three anonymous operators (TK, NU, CT).

```{r}
head(titration)
```

First, we need to compute the log dilution.

```{r}
titration$log_dil <- -log10(titration$dil)
```

The plot suggests that the tests give roughly similar ED50 values, but some data points are missing.  For example, Vial 3 by TK is missing a count at `log_dil == 6` and Vial 1 by CT is missing counts for `log_dil >= 6`.

```{r}
#| fig-alt: "Nine plots showing titration data."
ggplot(titration, aes(x = log_dil, y = positive)) +
  geom_point() +
  facet_grid(Vial ~ Operator) +
  geom_line()
```

## Example 1

We can estimate ED50 by aggregating the data from all nine tests for each of the three methods.

```{r}
#| message: false
print(
  SpearKarb(titration, positive + total ~ log_dil),
  verbose = FALSE
)
```

## Example 2

We can also check for differences between operators:

```{r}
#| message: false
print(
  SpearKarb(titration, positive + total ~ log_dil | Operator),
  verbose = FALSE
)
```

## Example 3

Or by vial:

```{r}
#| message: false
print(
  SpearKarb(titration, positive + total ~ log_dil | Vial),
  verbose = FALSE
)
```

## Example 4

Or by operator and vial.

```{r}
#| message: false
print(
  SpearKarb(titration, positive + total ~ log_dil | Operator + Vial),
  verbose = FALSE
)
```

# Multiple Tests

A helper function `skrmdb.all()` estimates ED50 using all three methods provided by this package.  This function is called in the same way that `DragBehr()`, `ReedMuench()`, and `SpearKarb()` are called, and the output is similar.

```{r}
#| message: false
print(
  skrmdb.all(titration, positive + total ~ log_dil | Operator + Vial),
  verbose = FALSE
)
```


# Output

By default, the print statement for a `skrmdb` object gives an interpretation of the columns in the table.  This should help explain any ways in which the data deviate from the assumptions made by the methods (See the next section).

```{r}
#| message: false
DragBehr(titration, positive + total ~ log_dil | Operator + Vial)
```

# Method Assumptions

Each of these methods was designed under the following assumptions:

1.  The group size $n_i$ is constant.
1.  The log dilutions $\mathbf{x} = \{x_1, x_2, \dots, x_k\}$ form an increasing sequence.
1.  The sequence $\mathbf{x}$ is arithmetic.  That is, there is some number $\Delta$ such that $x_k = (k - 1) \times \Delta$ for all $k$.
1.  The count ratio sequence $\mathbf{r} = \{r_1, r_2, \dots, r_k\}$, defined by $r_i = y_i / n_i$, trends from smaller to larger.
1.  Preferably, the count ratio sequence is monotonic, that is $r_i \leq r_{i+1}$ for all $i$.
1.  The log dilution sequence brackets ED50. That is, there is an $i$ such that $x_i \leq \text{ED50} \leq x_{i+1}$. Note: This is **not** equivalent to saying that there is an $i$ such that $r_i \leq 0.5 \leq r_{i+1}$.  However, if it is the case that $0.5 < r_1$ or $r_k < 0.5$, then the log dilution sequence will not bracket ED50.

The functions in this package will automatically sort the data by `x` and then determine whether $y_i/n_i$ or $1-y_i/n_i$ better meets the criterion that the response sequence be increasing. The selected sequence will be used to estimate ED50. In the case of noisy data, it may be best to gather more data or optimize the experiment before using any of these methods.

To illustrate this, we create some data where the number dead decreases as dilution increases, and so the number alive is increases.

```{r}
dead  <- c(10, 10, 8, 5, 3, 0)
total <- rep(10, 6)
alive <- total - dead
dil   <- 1:6
```

Given the assumptions that the methods make, it is appropriate to estimate ED50 from `alive + total ~ dil`, but we see that no matter how we enter the data, the same ED50 is reported.

```{r}
#| message: false
SpearKarb(alive + total ~ dil)$ed
SpearKarb(dead + total ~ dil)$ed
```

We can turn off the auto-sort feature, which will still give us a correct estimate of ED50 for `alive + total ~ dil`, but and incorrect estimate for `dead + total ~ dil`.

```{r}
#| message: false
SpearKarb(alive + total ~ dil, autosort = FALSE)$ed
SpearKarb(dead + total ~ dil, autosort = FALSE)$ed
```


# Accessor Functions

Finally, there are three accessor functions to help retrieve information from the results of `DragBehr()`, `ReedMuench()`, and `SpearKarb()`.

Using the titration data set again:

```{r}
res <- SpearKarb(titration, positive + total ~ log_dil)
```

We can retrieve the estimated ED50:  (**Note:** `getED50()` only works on `skrmdb` objects,  which are returned by `DragBehr()`, `ReedMuench()`, and `SpearKarb()`, not on `skrmdb.all` objects.)

```{r}
getED50(res)
```

The variance:

```{r}
getVar(res)
```

The data which were used:

```{r}
getData(res)
```

The results table:

```{r}
getResults(res)
```
